\documentclass{ximera}

\input{../../preamble.tex}

\title{Injective Linear Transformations and Dimension}

\begin{document}
\begin{abstract}
  Injective linear transformations cannot reduce dimension.
\end{abstract}
\maketitle


\begin{theorem}[Injective Linear Transformations and Dimension]
\label{theorem:ILTD}

Suppose that $\ltdefn{T}{U}{V}$ is an injective linear transformation.  Then $\dimension{U}\leq\dimension{V}$.

\begin{proof}
Suppose to the contrary that $m=\dimension{U}>\dimension{V}=t$.  Let $B$ be  a basis of $U$, which will then contain $m$ vectors.  Apply $T$ to each element of $B$ to form a set $C$ that is a subset of $V$.  By \ref{theorem:ILTB}, $C$ is linearly independent and therefore must contain $m$ distinct vectors.  So we have found a set of $m$ linearly independent vectors in $V$, a vector space of dimension $t$, with $m>t$.  However, this contradicts \ref{theorem:G}, so our assumption is false and $\dimension{U}\leq\dimension{V}$.

\end{proof}
\end{theorem}

\begin{example}

Consider the linear transformation 
\[\ltdefn{T}{M_{23}}{\complex{4}},\quad
\lteval{T}{\begin{bmatrix}a&b&c\\d&e&f\end{bmatrix}}=
\colvector{a+2b+12c-3d+e+6f\\2a-b-c+d-11f\\a+b+7c+2d+e-3f\\a+2b+12c+5e-5f}
\]


Since $\dimension{M_{23}}=6>4=\dimension{\complex{4}}$, 
\begin{multipleChoice}
\choice{$T$ must be injective}
\choice[correct]{$T$ cannot be injective}
\end{multipleChoice}
\begin{feedback}[correct]
Exactly!  If it were, then $T$ would violate \ref{theorem:ILTD}.
\end{feedback}

\end{example}

Notice that the previous example made no use of the actual formula defining the function.  Merely a comparison of the dimensions of the domain and codomain are enough to conclude that the linear transformation is not injective.  \ref{archetype:M} and \ref{archetype:N} are two more examples of linear transformations that have ``big'' domains and ``small'' codomains, resulting in ``collisions'' of outputs and thus are non-injective linear transformations.
\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
